Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-26 manuscript.rmd
Made concise to fit into word count limit

Section  | Word Count
------------- | -------------
Summary | 197
Main Text  | 6640
Introduction | 892
Methods | 3455
Results | 871
Discussion | 1422
Keywords | 6

\

Item | Count
------------- | -------------
Figures in color | 8
Tables | 2
Supplementary Material |  1 file

@@

Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance[@McDowell.2008]. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

@@

\
\      We tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

\
Table \@ref(tab:hypo) here.

# Materials and Methods
\
\   We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We defined species ERD as the depth at which soil water potential best explained species-specific 5-yearly mean growth rates over 25 years. To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model. The ERD model incorporated the impact of atmospheric drought on growth and was constrained with species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing deuterium to hydrogen content data for a subset of six tree species as an independent observation of ERD. We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets. Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.  
\
Fig. \@ref(fig:schematic) here.

@@

We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020]. Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but at a maximum only 6.3% *trees* in the canopy, and only 9.7% of the total forest crown area, is deciduous [@Condit:2000].

\
\     In the 50 ha plot, all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

\begin{equation}
\begin{split}
\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},i} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
\end{split}
(\#eq:water-balance)
\end{equation}
where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (typically, 1 hour). The soil profile is discretized into 15 exponentially distributed soil layers with depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.

@@

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq. \@ref(eq:water-balance)), (ii) stream discharge $q_\textrm{run}$ from the Conrad catchment, comparable to $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) comparable to $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).

@@

\   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized RMSE between observations and simulations for VWC at the depth of 1 m ($\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration. We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).

@@

We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.

@@

To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter to predict maximum diurnal ($K_\textrm{leaf}$), which in turn controls growth. $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential. This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimation ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close [@Grossiord.2020]. Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
(\#eq:M-chosen)
\end{equation}
where n~t~ be the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term. See SI for equations for all the alternate models tested.

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood (or, the minimum AIC score), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $R$^2 >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big\{z,h\,\Big|\, \Big(\textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation}
where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers below 0.21 m, averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.

@@

For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun. Calculating individual tree growth rates across six censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size [see SE], we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model, for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).

@@

We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods) described as:

\begin{equation}
K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + \epsilon_{s}
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters, $\epsilon_{s}$ an error term. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@

Table \@ref(tab:symbols) here.

>>>>

Section  | Word Count
------------- | -------------
Summary | 197
Main Text  | 6500
Introduction | 890
Methods | 3294
Results | 804
Discussion | 1512
Keywords | 6

\
\
Item | Count
------------- | -------------
Figures in color | 8
Tables | 2
Supplementary Material |  1 file

@@

Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones [@McDowell.2018], leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008]. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

@@

\
\      We tested an overarching hypothesis that inter-specific variation in rooting depths enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates in droughts due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

# Materials and Methods
\
\   We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)), defined as the depth at which soil water potential best explained species-specific growth dynamics. The ERD model incorporated the impact of atmospheric drought on growth and was constrained with species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest. To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model. We evaluated ERD against existing deuterium to hydrogen content data for a subset of six tree species as an independent observation of ERD. We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets. Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality.

@@

We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, hydrology and plant functional ecology, including data for stable water isotopes on trees and soils [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020]. Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but at a maximum only 6.3% *trees* in the canopy are deciduous, amounting to only 9.7% of the total forest crown area [@Condit:2000].

\
\     In the 50 ha plot, all stems above ten mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a two ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

\begin{equation}
\begin{split}
\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},i} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
\end{split}
(\#eq:water-balance)
\end{equation}
where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr). The soil profile is discretized into 15 exponentially distributed soil layers by depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.

@@

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted GWC to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).

@@

\   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration. We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near vertical probe (see [SI](#SIData.VWC)).

@@

We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme soil droughts were identified by depth $z$ as days that fell below the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.

@@

To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K_\textrm{leaf}$, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions). $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential. This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020]. Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
(\#eq:M-chosen)
\end{equation}
where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term. See SI for equations for all the alternate models tested.

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth. Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big\{z,h\,\Big|\, \Big(\textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation}
where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers below 0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.

@@

For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun. Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).

@@

We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods) described as:

\begin{equation}
K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + \epsilon_{s}
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters, $\epsilon_{s}$ an error term. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@
